Developing effective theories for single and bilayer graphene using density matrix downfolding

### 2019-09-09 draft.tex
updated preliminary work: MSJNO, CuO intro, fitting and results

The training data required for downfolding are extracted from the projected low-energy states via a mixed estimator.
Due to the fixed-node approximation, the final distribution returned by the projection is $f(R) = \Psi_T(R) \Psi_\text{DMC}$, and Monte Carlo estimates of operator expectation values correspond to the mixed estimation $\frac{\langle \Psi_T | \hat{O} | \Psi_\text{DMC} \rangle}{\langle \Psi_T | \Psi_\text{DMC}\rangle}$. 
This does not equal the target expectation value $\frac{\langle \Psi_\text{DMC} | \hat{O} | \Psi_\text{DMC} \rangle}{\langle \Psi_\text{DMC} | \Psi_\text{DMC}\rangle}$, resulting in a mixed estimator error.
Details on alleviating this error in DMD and evaluation of expectation values for 1- and 2-body second quantized operators can be found in the literature \cite{Wagner2015, Wagner2017}.
\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with our code QWalk \cite{Wagner2009} and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{doi:10.1063/1.5052906}.
The MSJ+NO trial wave functions take the form
\begin{equation}
\Psi=e^{J(\vec{\alpha})} \sum_I e^{\hat{\Theta}(\vec{\theta_I})} C_I |D_I (\{ \phi\})\rangle
\end{equation}
where the determinants $|D_I\rangle$ are composed of single particle orbitals from a self-consistent field calculation, $e^J$ is a spin-independent Jastrow factor, $e^{\hat{\Theta}}$ is a spin-dependent orbital rotation matrix, and the wave function has variable parameters $\vec{\alpha}, \vec{\theta}, \vec{C}$ which can be optimized.

We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in both variational Monte Carlo (VMC) and FN-DMC calculations. 
We found the average improvement in variational energy when using optimized orthogonal determinants is $\langle E_\text{MSJ} - E_\text{MSJ+O} \rangle = $ 0.32 eV with a standard deviation of 0.08 eV, with an additional improvement from the non-orthogonal determinants of 0.031 eV with a standard deviation of 0.021 eV.
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was 0.14 eV (standard deviation 0.03 eV) with an additional reduction of 0.032 eV (standard deviation 0.019 eV) when using the MSJ+NO trial wave function.

While the benefit using the MSJ+NO trial wave function seems small, we saw that the FN-DMC energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the FN-DMC energy using an MSJ+O trial function with 55 determinants.
Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
Our results indicated that using non-orthogonal determinants may lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
As a first step towards developing accurate models for extended systems like solids using DMD, I constructed a many-body effective model for the CuO molecule by downfolding \eqref{Hab} under a Born-Oppenheimer approximation with fixed bond length $r_e$ = 1.725\r{A}.
In the following four sections I will detail the steps I took towards developing a model for the CuO molecule using DMD.
First, I discuss how low-energy states were sampled for the model regression.
This is followed by details of descriptor selection and single particle basis construction.
With the sampled data and candidate descriptors in hand, the feature selection and model fitting procedures are described.
I will conclude by comparing the solutions of the model Hamiltonian to experiment and previous theoretical calculations.

@@

\vspace{-10pt}
\paragraph{Fitting the model}
A direct attempt at fitting all eleven terms in \eqref{eq:models} leads to a high goodness of fit, R$^2$ = 0.99, but unexpectedly large coefficients for the fit model, like an enormous occupation energy gap of $\epsilon_{2p_\pi} - \epsilon_{3d_\delta} = $ 5.87 eV, which is 2 $-$ 3 eV in experiment and our sampled data.
This inflation of model parameters was found to be associated with multicollinearity among the candidate descriptors, quantified by a large condition number (CN) of 1470.

A principal component regression (PCR) was used to alleviate this multicollinearity and fit the effective theory \cite{10.2307/2348005}.
PCR can be broken into three parts: conducting a principal component analysis (PCA) and selecting candidate principal components to work with (PCs), fitting a linear model to these PCs, and then inverting the coefficients back to the original descriptors.
If the subset of PCs are chosen appropriately, the final regressed model will have a similar goodness of fit to one fit on the original data matrix but without the consequence of serious multicollinearity.

From the PCA, the nine highest ranked PCs were selected to use in the PCR.
In Figure \eqref{fig:cuo_regr}b we plot both the cumulative explained variance ratio (CEVR) and CN of a matrix composed of the $N_\text{comp}$ highest ranked PCs varying from one to eleven.
The CEVR indicates how much of the variation in the descriptor values a subset of PCs capture, and the CN quantifies the multicollinearity among these PCs.
The objective in selecting PCs is to the maximize the prior while simultaneously minimizing the latter, ensuring a high goodness of fit with little to no multicollinearity.
Nine principal components best satisfies the objective above with a CEVR 0.9995 and CN of 257.

The model fit to these nine PCs maintains an excellent goodness of fit without serious multicollinearity.
When trained on all the sampled data this model yields an $R^2$ = 0.98 with a CN of 257. 
An inverse PCA transformation was used to map the fit coefficients corresponding to the PC descriptors back into the original eleven coefficients of \eqref{eq:models}.
These coefficients can be seen in Figure \ref{fig:cuo_regr}c with single standard deviation bootstrap error estimates.

>>>>

The training data required for downfolding are extracted from the projected low-energy states via a mixed estimator.
Due to the fixed-node approximation, the final distribution returned by the projection is $f(R) = \Psi_T(R) \Psi_\text{DMC}$, and Monte Carlo estimates of operator expectation values correspond to the mixed estimation $\frac{\langle \Psi_T | \hat{O} | \Psi_\text{DMC} \rangle}{\langle \Psi_T | \Psi_\text{DMC}\rangle}$. 
This does not equal the target expectation value $\frac{\langle \Psi_\text{DMC} | \hat{O} | \Psi_\text{DMC} \rangle}{\langle \Psi_\text{DMC} | \Psi_\text{DMC}\rangle}$, resulting in a mixed estimator error.
Details on alleviating this error in DMD and evaluation of expectation values for 1- and 2-body second quantized operators can be found in the literature \cite{Wagner2015, Wagner2017}.
\section{Preliminary Work}
\subsection{Non-orthogonal determinants in FN-DMC trial wave functions}
In order to become acquainted with QMC algorithms I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized non-orthogonal determinants (MSJ+NO) in FN-DMC \cite{doi:10.1063/1.5052906}.
The MSJ+NO trial wave functions take the form
\begin{equation}
\Psi=e^{J(\vec{\alpha})} \sum_I e^{\hat{\Theta}(\vec{\theta_I})} C_I |D_I (\{ \phi\})\rangle
\end{equation}
where the determinants $|D_I\rangle$ are composed of single particle orbitals from a self-consistent field calculation, $e^J$ is a spin-independent Jastrow factor, and $e^{\hat{\Theta}}$ is a spin-dependent orbital rotation matrix.
The wave function has variable parameters $\vec{\alpha}, \vec{\theta}, \vec{C}$ which can be optimized.

I assessed the efficiency and compactness of the new function by calculating the ground state energy of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with orthogonal determinants (MSJ+O). 
The workflow involved constructing the un-optimized trial wave functions, optimizing the parameters using an energy optimization method \cite{Toulouse2007}, and finally using the optimized trial functions in FN-DMC calculations. 
The average improvement in the FN-DMC energy with the MSJ+O trial wave functions was $\langle E_{MSJ} - E_{MSJ,O}\rangle =$ 0.14eV with an additional reduction of 0.032 eV when using the MSJ+NO trial wave function.
While the benefit using the MSJ+NO trial wave function seems small, the energy calculated using an MSJ+NO trial function with only 24 determinants was lower than the energy using an MSJ+O trial 
function with 55 determinants.

Further, the FN-DMC charge density calculated using MSJ+NO trial functions had stronger bonding character than when using MSJ+O trial functions, a reasonable result as introducing correlations into trial functions allows for electrons to avoid each other while still occupying the same bonding region. 
The results indicated that using non-orthogonal determinants may lead to more compact trial wave functions for small molecules.

\subsection{Effective theory for CuO molecule using DMD}
To become acquainted with DMD, I constructed a many-body effective model for the CuO molecule with fixed bond length $r_e$ = 1.725\r{A}.
In the following three sections I will detail the steps I took towards developing a model for the CuO molecule using DMD, followed by a comparison of model spectra to experiment.

@@

\vspace{-10pt}
\paragraph{Fitting the model}
A direct attempt at fitting all eleven terms in \eqref{eq:models} leads to a high goodness of fit, R$^2$ = 0.99, but unexpectedly large coefficients for the fit model.
This is seen in an enormous occupation energy gap of $\epsilon_{2p_\pi} - \epsilon_{3d_\delta} = $ 5.87 eV, which is 2 $-$ 3 eV in experiment and our sampled data.
This inflation of model parameters was found to be associated with multicollinearity among the candidate descriptors, quantified by a large condition number (CN) of 1470.

A principal components analysis (PCA) on the descriptors illustrates the multicollinearity in the data set.
In Figure \eqref{fig:cuo_regr}b I show the cumulative explained variance ratio (CEVR) and CN of a matrix composed of the $N_\text{comp}$ highest ranked PCs varying from one to eleven.
The CEVR indicates how much of the variation in the descriptor values a subset of PCs capture, and the CN quantifies the multicollinearity among these PCs.
The sharp decrease in the condition number and sustained high CEVR up to nine principal components indicates the presence of only nine linearly independent variations within the descriptor set.

To reduce multicollinearity while maintaining a high goodness of fit, I fit a model to these nine PCs.
When trained on all the sampled data this model yields an $R^2$ = 0.98 with a CN of 257. 
An inverse PCA transformation was used to map the fit coefficients corresponding to the PC descriptors back into the original eleven coefficients of \eqref{eq:models}.
These coefficients can be seen in Figure \ref{fig:cuo_regr}c with single standard deviation bootstrap error estimates.
The procedure carried out here is well established and called principal components regression (PCR) \cite{10.2307/2348005}.

### 2019-09-09 reoutline.tex
updated preliminary work: MSJNO, CuO intro, fitting and results

>>>>

\section{Non-orthogonal determinants in FN-DMC trial wave functions}
\begin{enumerate}
\item In order to become acquainted with our code QWalk and QMC algorithms in general I worked on implementing and testing multi-Slater-Jastrow trial functions with optimized nonorthogonal determinants (MSJ+NO) in FN-DMC.

\item We assessed the efficiency and compactness of this new trial function by calculating the ground state energy and single particle densities of a C$_2$ molecule using FN-DMC and comparing to the results when using multi-Slater-Jastrow trial functions with optimized orthogonal determinant trial functions (MSJ+O).

\item Our results indicated that using non-orthogonal determinants may
lead to more compact multi-Slater-Jastrow trial wave functions for small molecules.
\end{enumerate}

\section{Effective theory for CuO molecule using DMD}
\begin{enumerate}
\item To become acquainted with DMD, I constructed a many-body effective model for the CuO molecule with fixed bond length r$_e$ = 1.725A.

%sampling low energy states
\item Explain why MSJ was chosen (Probably the hardest one)

\item How determinants selected

\item How coefficients selected

\item How Jastrow optimized and FN-DMC done

%candidate descriptors 
\item Physical intuition and symmetry

\item Statistics

\item Final set and basis elements

%Fitting the model 
\item A direct attempt at fitting all eleven terms in \textbf{eqref} leads to a high goodness of fit, R$^2$ = 0.99, but unexpectedly large coefficients for the fit model.

\item A principal components analysis (PCA) on the descriptors illustrates the multicollinearity in the data set.

\item To reduce multicollinearity while maintaining a high goodness of fit, I fit a model to these nine PCs.

%Solutions of the model 
\item The model is solved by exact diagonalization, and the solutions agree well with experimental spectra and state assignments.

\item The model states also agree with recent theoretical calculations like UKS and CCSD(T).

\end{enumerate}
